\large Differentiable Cosmological Simulation with the Adjoint Method \vspace{0.3em}

### 2023-11-02 docs/papers/adjoint/response.txt

We thank the referee for the careful review, which really helped improve
the manuscript. And YL apologize for the delay in this response, due to
mostly job related reasons. We address each point below.

> The authors present a novel framework for differentiable cosmological
> simulations based on the adjoint method, implemented in the code
> package “pmwd”. A great advantage of their method is that the memory
> usage is independent of the number of timesteps performed; hence, this
> paper is pioneering in the development of differentiable simulations
> with a large number of timesteps. The authors achieve this by
> computing the gradients of some objective function w.r.t. the
> cosmological parameters and the initial white noise field via the
> adjoint equations, unlike previous differentiable simulations that
> backpropagate the gradients through all time steps of the simulation
> with automatic differentiation. Using the adjoint method for
> cosmological N-body simulations is an excellent idea, and while the
> authors only consider a simple toy example as the only application in
> this work, which does not reveal the full potential of the method, I
> am certain that their framework will find use in many applications in
> the context of forward modeling.
>
> The paper is generally well written and mostly well structured;
> however, the presentation of the adjoint method could still be
> improved, so I would suggest the authors to restructure that section.
> Altogether, I am confident that the manuscript will eventually merit
> publication in ApJS, but I would like to kindly ask the authors to
> address my detailed comments below.
>
> General comment:
> • For readers unfamiliar with the adjoint method, the intuitive
> meaning of the adjoint variables is somewhat opaque, and the adjoint
> equations are not well motivated in the main text. When the adjoint
> equation is introduced above l.239, it is unclear how the adjoint
> variables xi and pi relate to x, p, and to the objective function J,
> and why they are needed in the first place. I would suggest to state
> the initial conditions for the adjoint variables xi and pi (i.e.
> ∂J/∂x_n and ∂J/∂p_n) already before the detailed equations for their
> evolution. In this way, it immediately becomes clear that xi and pi
> are needed for tracking the gradient of the objective function J
> w.r.t. to x and p backwards from the final time to the initial
> conditions. Also, it might be more pedagogical to start Sec. 2.5 with
> Eq. (20) to emphasize that your aim is to compute the total derivative
> dJ/dθ, which can be expressed in terms of auxiliary variables (namely
> the adjoint variables) and only then provide the detailed equations
> (18) that the adjoint variables satisfy. It should be highlighted
> that, crucially, the definition of the adjoint variables depends on
> the objective J (and changing the objective requires re-computing the
> backward pass of the simulation with a different set of adjoint
> variables).

We have expanded the introduction section extensively with a high-level
introduction of the adjoint method, including some intuition on the
adjoint variables, the adjoint equations, their initial conditions, and
the gradients on parameters. We also think following through the
pedagogical development in the previous App. B is probably the best way
to get a better understanding of the method, since the formalism there
is both simpler and more generic than its specific application in Sec.
2.5. So we have moved that appendix before the previous Sec. 2.5 and
merge them into a new Sec. 3.

> Comments:
> • Introduction: the authors emphasize the use of automatic
> differentiation in their framework (as compared to analytic
> derivatives used in earlier works). For completeness, could the
> authors please add a brief explanation (1-2 sentences) of how AD works
> and how it compares to analytic and numerical differentiation?

Explanations of the analytic and automatic differentiations are added.

> • l.67: Could the authors please define the “discretize-then-optimize”
> approach (in comparison to the “optimize-then-discretize” approach)?

Explanations of both approaches added.

> • l.192: When the authors mention that FastPM chooses G_D and G_K
> “according to the ZA growth history”, they should make the connection
> to the 2LPT equation (2) to emphasize that the ZA is actually 1LPT.

Connection added.

> • Sec. 2.4: The observations “on our past light cone” seem somewhat of
> an overkill because the authors only consider a toy example in the
> results section, where the observable is given by the density field in
> the simulation box at final time. But I guess the authors are planning
> on using their framework in more complex settings in future work, so
> this is probably fine.

Indeed, we have derived the formalism for the light cone in Sec. 2.4 and
Sec. 3.4-3.6. And we have implemented snapshots as pmwd "observables"
and hopefully will have on-the-fly light cones and even post-Born ray
tracing available very soon.

> • Sec. 2.5: The authors should clearly state the meaning of the
> adjoint variables when they appear for the first time (see my comment
> above). In fact, this should already be mentioned in the introduction.
> I think it would already provide a lot of intuition for the somewhat
> abstract notion of adjoint variables and equations if the authors
> explained earlier in the paper that the adjoint variables are
> initialized as the gradient of the objective function J w.r.t. the
> state variables (x, p), i.e. the first two expressions in Eq. (19).
> Note that the expression that the adjoint method aims to compute,
> namely dJ/dθ, only appears at the very end of Sec. 2.5, rather than at
> the beginning.

Clarified in the introduction.

> • l.239: “...and left the ∂J/∂z_i term to the O operator”: based on
> the main text alone, it is unclear which ∂J/∂z_i term the authors are
> referring to here.

Some intuition added.

> • l.262: “Also note that θ can stand for both itself and ω, which we
> have omitted for brevity”: Given that the authors emphasize the
> distinction between θ and ω in Fig. 1, I would suggest changing the
> notation to e.g. θ* := (θ, ω) for clarity and consistency.

Equation for ω added explicitly.

> • l.280: “Because GPUs are inherently parallel devices, they can
> output different results for identical inputs.” This is not clear to
> me – where does the non- determinism come from specifically? I was
> under the impression that at least in the newer versions of Jax (or
> when setting TF_CUDNN_DETERMINISTIC) Jax is deterministic even on a
> GPU, see the discussion here:
> https://github.com/google/jax/issues/565. Could the authors please
> comment on this?

The non-determinism result for example from the different orders in
reduction operations among multiple processors. Enforcing determinism on
GPUs hurt the performance of those opertators. Some later issue
(https://github.com/google/jax/issues/13672) mentioned the flag to set
is now XLA_FLAGS='--xla_gpu_deterministic_ops=true'. A test run on a
problem 32768 times smaller than that in Fig. 2 (with 16^3 particles)
takes 10s to finish, so the deterministic run seems be more than 10^5
times slower (also accounting for the logN factor).

> • l.296: “... in practice the reconstructed trajectory can differ from
> the forward one due to loss of small-scale information”. What exactly
> do the authors mean by “small-scale information” here? As the authors
> state, the N-body equations themselves are time- reversible, so any
> non-reversibility should be caused by numerical errors, and the
> authors should be more specific as to what exactly causes the
> non-reversibility.

Clarified following the suggestion.

> • Tables 1 and 2: since the number of particles seems to be fixed at
> 384^3 in the reproducibility and forward-vs-backward experiments, the
> authors vary the particle mass by effectively varying the box size if
> I am not mistaken. For completeness, it would be good to state the box
> size that belongs to each particle mass.

Clarified following the suggestion. The pmwd API actually ask user to
provide the Lagrangian particle spacing first, which is directly related
to the particle mass. This allows us to specify a non-cubic box more
easily.

> • Table 2: The relative difference in the backwards-in-time
> displacement and velocity field towards the 2LPT ICs can be quite
> large, e.g. 5.2% / 7.1% for the first case. It would be interesting to
> mention how this error compares to the magnitude of the 2LPT
> displacement component s^(2) in Eq. (3). For example, if one starts
> with 2LPT at z = z_ini, evolves the system forward in time to z = 0,
> then goes backwards in time again and finally ends up with an error on
> the same order of magnitude as the 2LPT correction to the Zel’dovich
> term or even larger at z = z_ini, it might be questionable if the
> accuracy of the backward evolution is sufficient.

An interesting point. Since high-order LPTs affect the forward model
accuracy of the late time observables, and reversibility affects the
gradient accuracy on the simulation inputs, it seems to me that they are
orthogonal effects and a fair comparison is more complicated than one at
the 2LPT level. Ideally, we want both accuracy. And we are implementing
the 3LPT which can directly help the former, and indirectly the latter
too. We plan to test their impacts on the inference in future works.

> • Caption of Table 2: “Their smallness contributes partly to the fact
> that the quoted relative differences here being much greater than
> those in Table 1.” If this is indeed the problem, the authors could
> modify / expand their time reversibility experiment and evolve their
> backwards-in-time solution at the initial time once again forward in
> time and compare the difference at the final time (i.e. forward vs.
> forward-backward- forward) to make Tables 1 and 2 more comparable.

A fairer and more direct comparison is actually already in Fig. 3, where
the adjoint gradients suffer from both the reproducibility and the
reversibility, while the AD results only from the former. We have edited
the caption and the text to highlight this.

> • l.307: in the AD vs. adjoint gradient comparison, it is unclear to
> me which gradients are compared. From the text, I understand the
> gradients are w.r.t. the white noise field and the cosmological
> parameters, but what is the objective function J for this comparison?
> Please add this information.

Added in the caption of Fig. 3.

> • l.314: “we perform both adjoint and AD runs for 8 times”: I assume
> these 8 runs are done with the same cosmological parameters, but with
> different white noise fields – could the authors please comment on
> this and add this information to their manuscript?

Actually we repeated 8 times with the same cosmology and the same white
noise modes. This allows us to compare not only just the adjoint and AD
gradients, but also the adjoint or AD with itself, i.e., the
reproducibility of either gradients. We have clarified this in the text.

> • l.323: “possible breaking of order- or time-invariance in AD”: what
> exactly do the authors mean by this (in particular by “order
> invariance in AD”)?

This is related to the last point and should hopefully be clearer now.
The 8 AD runs give slightly different results that correlate with their
order, so that their asymmetric differences has asymmetric tails in the
histogram. However, with a new combination of JAX, CUDA and GPU, this
problem has disappeared, which we have further verified by increasing
the repetition to 64 times. So we have removed this comment.

> • Fig. 3: Could the authors please provide a description of the
> red-blue graphic and also add a color bar?

Description of the upper panel and color bar are added.

> • Fig. 4: The toy example is very illustrative; however, it would be
> interesting to also see the optimized ICs (i.e., the ICs for which the
> forward simulation leads to the “pmwd” pattern) – I encourage the
> authors to add them to the figure.

We found that the variance of the modes became bigger than 1 (that of
the standard normal white noise) after the optimization. We plot the
projected mean and standard deviation of these modes confirming that
trend. In addition, the mean figures show some level of spatial
correlation not present in the white noise modes. We have added these
figures to the repository
(https://github.com/eelregit/pmwd/tree/master/docs/papers/adjoint
optim_modes_{mean,std}_{0,10,100,1000}.pdf) and summarize the finding in
the text.

> Minor comments:
> • l.40: “rapid advances ... opens” → “open”
> • l.53: “..., thus requires saving the states...” → “..., which
>   requires saving the states...”?
> • l.56: “much smaller than” → “much less than”?
> • l.62: “these variables evolves” → “evolve”
> • l.192: “thereby improves” → “thereby improving”
> • l.245: “but enter” → “but enters”
> • l.256: “we can accumulate objective gradient following” → “the
>   objective gradient”?
> • l.271: it seems as if an equal sign “=” is missing here: “θ(As = 2 ×
>   10−9” → “θ = (As = ...)
> • l.293: “To save memory, trajectory of the model state” → “the
>   trajectory”?
> • Caption of Table 2: “... that the quoted relative differences here
>   being much greater than those in Table 1.” → “...are much greater”?
> • l.313: “To compensate for good statistics” → this sounds somewhat
>   odd to me (shouldn’t it rather be “to compensate for bad
>   statistics”?), maybe reformulate
> • l.318: “agree very well, and so are their gradients” → “so do their
>   gradients”?
> • l.441: “...using the facts that d(M^{−1}) = −M^{−1} dM M^{−1}”: it
>   took me a second to realize that M stands for an arbitrary matrix
>   here – maybe mention this briefly
> • l.444: “when one optimize” → “optimizes”

Corrected following the referee's suggestions.

We have also updated the results, especially Fig. 5, using an H100 GPU
instead of the previous A100.

> Data/Software Review:

We thank the data editor for suggestions. And YL apologize for the delay
in this response, due to mostly job related reasons.

> I. It is not clear the state of the pmwd JOSS submission. Is there a
> submission/review link started at JOSS? Do you wish this to be a joint
> AAS+JOSS review article [1]? This information would be useful both for
> the editing of the manuscript and for the eventual reader to
> understand how pmwd relates to the ApJS article. Please reply to this
> in the reply to science editor indicating if you want these journal
> articles related.
> [1] https://blog.joss.theoj.org/2018/12/a-new-collaboration-with-aas-publishing

I have not submitted to JOSS yet as there are still some unit testing,
documentation, and cleaning to be done. This is also my first AAS+JOSS
attempt. Should the two reviewing processes be coordinated? I will do my
best to submit to JOSS as soon as possible.

> II. I would suggest depositing the specific scripts related to this
> ApJS submission into a DOI-issuing repository and linking the text to
> that DOI -- in addition to, but not replacing the current GitHub
> script link. More information on this can be found in our tutorial [2]
> [2] https://github.com/AASJournals/Tutorials/tree/master/Repositories

To accommodate some new features, we plan to change the API slightly
in the following weeks. I will then modify the scripts accordingly for
the permanent repository and provide the DOI.

### 2023-11-09 docs/papers/adjoint/adjoint.tex

\documentclass[modern, trackchanges, dvipsnames]{aastex631}
  \urlstyle{sf}

@@

\newcommand{\GPU}{NVIDIA A100 80GB SXM4}

\newcommand{\HL}[1]{\textcolor{Bittersweet}{#1}}

@@

Rapid advances in deep learning have brought not only myriad powerful
neural networks, but also breakthroughs that benefit established
scientific research.
In particular, automatic differentiation (AD) tools and computational
accelerators like GPUs have facilitated forward modeling of the Universe
with differentiable simulations.
Current differentiable cosmological simulations are limited by memory,
thus are subject to a trade-off between time and space/mass resolution.
They typically integrate for only tens of time steps, unlike the
standard non-differentiable simulations.
We present a new approach free of such constraints, using the adjoint
method and reverse time integration.
It enables larger and more accurate forward modeling, and will improve
gradient based optimization and inference.
We implement it in a particle-mesh (PM) $N$-body library \pmwd\
(particle-mesh with derivatives).
Based on the powerful AD system \texttt{JAX}, \pmwd\ is fully
differentiable, and is highly performant on GPUs.

@@

Current established workflows of statistical inference from cosmological
datasets involve reducing cleaned data to summary statistics like the
power spectrum, and predicting these statistics using perturbation
theories, semi-analytic models, or simulation-calibrated emulators.
These can be suboptimal due to the limited model fidelity and the
risk of information loss in data compression.
Cosmological simulations \citep{HockneyEastwood1988, AnguloHahn2022} can
accurately predict structure formation even in the nonlinear regime at
the level of the fields.
Using simulations as forward models also naturally accounts for the
cross-correlation of different observables, and can easily incorporate
systematic errors.
This approach has been intractable due to the large computational costs
on conventional CPU clusters, but rapid advances in accelerator
technology like GPUs opens the possibility of simulation-based modeling
and inference \citep{CranmerEtAl2020}.
Furthermore, model differentiability enabled by AD libraries can
accelerate parameter constraint with gradient-based optimization and
inference.
A differentiable field-level forward model combining these two features
is able to constrain physical parameters together with the initial
conditions of the Universe.

The first differentiable cosmological simulations, such as BORG, ELUCID,
and BORG-PM \citep{BORG, ELUCID, BORG-PM}, were developed before the
advent of modern AD systems, and were based on implementations of
analytic derivatives \citep[see e.g.,][App.~D]{SeljakEtAl2017}).
Later codes including \texttt{FastPM} and \texttt{FlowPM} \citep{FastPM,
vmad, SeljakEtAl2017, FlowPM} compute gradients using AD engines, namely
\texttt{vmad} (written by the same authors) and \texttt{TensorFlow},
respectively.
Both analytic differentiation and AD backpropagate the gradients through
the whole history, thus requires saving the states at all time steps in
memory.
Therefore, they are subject to a trade-off between time and space/mass
resolution, and typically can integrate for only tens of time steps,
much smaller than the standard non-differentiable simulations.
As a result, they lose accuracy on small scales and in dense regions
where the time resolution is important, e.g., in weak lensing
\citep{MADLens}.

Alternatively, the adjoint method provides systematic ways of deriving
model gradients under constraints \citep{Pontryagin1962}, such as those
imposed by the $N$-body equations of motion in a simulated Universe.
It identifies a set of adjoint variables, dual to the state variables in
the simulation.
For time-dependent problems, these variables evolves backward in time by
a dual set of equations, known as the adjoint equations.
Their evolution depends on the states in the forward run, which can be
re-simulated with reverse time integration if the dynamics is
reversible, thereby dramatically reducing the memory cost
\citep{NeuralODE}.
Furthermore, we take the discretize-then-optimize approach
\citep[e.g.,][]{ANODE} to ensure gradients propagate backward along the
\emph{same} discrete trajectory as taken by the forward time
integration.
In brief, to compute the gradient we only need to evolve a simulation
forward, and then backward jointly with its dual adjoint equations.
We introduce the adjoint method for generic time-dependent problems in
both continuous and discrete cases in \autoref{app:adj}.

@@

field $\vomega$ indicate dependence on $\vtheta$ only, while double

@@

We describe different operators in \autoref{sec:method}: Boltzmann
solver (``Boltz'') and initial condition generator (``IC'') in

@@

\vspace{1em}
\section{Methods}
\label{sec:method}

We first review all components of the $N$-body simulation-based forward
model of the cosmological structure formation, and then introduce the
adjoint method for each part, and how they can be combined to
differentiate a whole simulation at low memory cost.

@@

The initial conditions of particles can be set perturbatively when
$|\vnabla \cdot \vs|$, the linear approximation of the density
fluctuation, is much smaller than 1.
We compute the initial displacements and momenta using the second order
Lagrangian perturbation theory \citep[2LPT,][]{BouchetEtAl1995}:
%
\begin{align}
\vs &= D_1 \vs^{(1)} + D_2 \vs^{(2)}, \nonumber\\
\vp &= a^2 \dot\vx = a^2 \dot\vs
  = a^2 H \bigl( D_1' \vs^{(1)} + D_2' \vs^{(2)} \bigr),
\label{ic}
\end{align}
%
where $\vp$ is the canonical momentum\footnote{We omit the particle mass
$m$ in the canonical momentum $\vp = m a^2 \dot\vx$ throughout for
brevity.} for canonical coordinate $\vx$.
The temporal and spatial dependences separate at each order: the $i$-th
order growth factor $D_i$ is only a function of scale factor $a$ (or
time $t$), and the $i$-th order displacement field $\vs^{(i)}$ depends
only on $\vq$.
We have also used two types of time derivatives, $\dot\Box \triangleq
\d\Box / \d t$ and $\Box' \triangleq \d\Box / \d \ln a$, related by
$\dot\Box = H \Box'$.

Both the first and second order displacements are potential flows,
%
\begin{equation}
\vs^{(1)} = - \vnabla_\vq \phi_\vs^{(1)},
\quad
\vs^{(2)} = - \vnabla_\vq \phi_\vs^{(2)},
\end{equation}
%
with the scalar potentials sourced by\footnote{Note the difference in
the sign of the source of $\phi_\vs^{(2)}$ from the usual LPT
convention.}
%
\begin{align}
\nabla_\vq^2 \phi_\vs^{(1)} &= \delta^{(1)}(\vq), \nonumber\\
%
\nabla_\vq^2 \phi_\vs^{(2)} &= {\textstyle \sum_{i<j}} \bigl(
  \phi_{\vs,ii}^{(1)}\phi_{\vs,jj}^{(1)}
  - \phi_{\vs,ij}^{(1)}\phi_{\vs,ji}^{(1)}
\bigr),
\end{align}
%
where $\phi_{\vs,ij} \triangleq \p^2\phi_\vs / \p q_i \p q_j$,
and $\delta^{(1)}$ is the linear order of the overdensity field,
$\delta$, related to the density field $\rho$ and mean matter density
$\bar\rho$ by $\delta \triangleq \rho / \bar\rho - 1$.

@@

Cosmological perturbation theory gives the linear power spectrum
as\footnote{Here the time dependence of the linear power, $\Plin(k, a) =
\Plin(k) D_1^2(a)$, has been left to \eqref{ic}}
%
\begin{equation}
\frac{k^3}{2\pi^2} \Plin(k)
= \frac{4}{25} \As
  \Bigl( \frac{k}{k_\mathrm{pivot}} \Bigr)^{\ns - 1}
  T^2(k) \frac{c^4 k^4}{\Omegam^2 H_0^4},
\label{Plin}
\end{equation}
%
where the shape of $\Plin$ is determined by the transfer function $T$,
solution to the linearized Einstein-Boltzmann equations \citep{CAMB,
CLASS}.
$T$ depends on the cosmological parameters,
%
\begin{equation*}
\vtheta = (
  \As, \ns, \Omegam, \Omegab, h, \cdots),
\end{equation*}
%
some of which already appear in \eqref{Plin}: $\As$ is the amplitude of
the primordial power spectrum defined at some fixed scale
$k_\mathrm{pivot}$; $\ns$ describes the shape of the primordial power
spectrum; $\Omegam$ is the total matter density parameter; $\Omegab$ is
the baryonic matter density parameter; and $H_0$ is the Hubble constant,
often parameterized by the dimensionless $h$ as $H_0 = 100 h \,
\mathrm{km}/\mathrm{s}/\Mpc$.
Other parameters may enter in extensions of the standard $\Lambda$ cold
dark matter ($\Lambda$CDM) cosmology.

@@

While our adjoint method is general, we employ the PM solver in \pmwd\
for efficiency, and leave the implementation of short-range forces to
future development.
With the PM method, we evaluate $\delta(\vx)$ on an auxiliary mesh by
\emph{scattering} particle masses to the nearest grid points.
We use the usual cloud-in-cell (CIC), or trilinear, interpolation
\citep{HockneyEastwood1988}, to compute the fractions of a particle at
$\vx'$ going to a grid point at $\vx$,
%
\begin{equation}
W(\vx, \vx') = \prod_{i=1}^3
  \max \Bigl( 1 - \frac{|x_i-x'_i|}{l_\mathrm{cell}}, 0 \Bigr),
\label{cic}
\end{equation}
%
where $l_\mathrm{cell}$ is the mesh cell size.

@@

$N$-body particles move by the following equations of motion
%
\begin{align}
\dot\vx &= \frac{\vp}{a^2}, \nonumber\\
\dot\vp &= -\vnabla\phi = - \frac{\vnabla\varphi}a.
\label{eom}
\end{align}
%
We use the \texttt{FastPM} time stepping \citep{FastPM}, designed to
reproduce the first order Lagrangian perturbation theory, also known as
the Zel'dovich approximation (hereafter ZA), in the linear regime.
We present a simplified derivation below.

@@

\texttt{FastPM} chooses $G_D$ and $G_K$ according to the ZA growth
history, thereby improves the accuracy on large scales and at early
times.
In ZA, the displacements are proportional to the linear growth factor,
$\vs \propto D_1$, which determines the time dependences of the momenta
and the accelerations by \eqref{eom}.
Therefore, we can set $G_D$ and $G_K$ in \eqref{eom_integ} to
%
\begin{align}
G_D &:= a^2 \dot D_1, \nonumber\\
G_K &:= a \dot G_D.
\label{G_ZA}
\end{align}
%
They are functions of $D_1$ and its derivatives, as given by
\eqref{G_eval}.
With these choices, the drift and kick factors, defined in
\eqref{eom_integ}, then become
%
\begin{align}
D(t_c, t_a, t_b) &=
  \frac{D_1(t_b) - D_1(t_a)}{G_D(t_c)}, \nonumber\\
%
K(t_c, t_a, t_b) &=
  \frac{G_D(t_b) - G_D(t_a)}{G_K(t_c)}.
\end{align}
%

@@

\vspace{1em}
\subsection{Adjoint Method}
\label{sec:adj}

>>>>

\documentclass[modern, dvipsnames]{aastex631}
  \urlstyle{sf}

@@

\newcommand{\GPU}{NVIDIA H100 PCIe}

\newcommand{\HL}[1]{\textcolor{Bittersweet}{\textbf{#1}}}

@@

Rapid advances in deep learning have brought not only myriad powerful
neural networks, but also breakthroughs that benefit established
scientific research.
In particular, automatic differentiation (AD) tools and computational
accelerators like GPUs have facilitated forward modeling of the Universe
with differentiable simulations.
Based on analytic or automatic backpropagation, current differentiable
cosmological simulations are limited by memory, and thus are subject to
a trade-off between time and space/mass resolution, usually sacrificing
both.
We present a new approach free of such constraints, using the adjoint
method and reverse time integration.
It enables larger and more accurate forward modeling at the field level,
and will improve gradient based optimization and inference.
We implement it in an open-source particle-mesh (PM) $N$-body library
\pmwd\ (particle-mesh with derivatives).
Based on the powerful AD system \texttt{JAX}, \pmwd\ is fully
differentiable, and is highly performant on GPUs.

@@

Current established workflows of statistical inference from cosmological
datasets involve reducing cleaned data to summary statistics like the
power spectrum, and predicting these statistics using perturbation
theories, semi-analytic models, or simulation-calibrated emulators.
These can be suboptimal due to the limited model fidelity and the
risk of information loss in data compression.
Cosmological simulations \citep{HockneyEastwood1988, AnguloHahn2022} can
accurately predict structure formation even in the nonlinear regime at
the level of the fields.
Using simulations as forward models also naturally accounts for the
cross-correlation of different observables, and can easily incorporate
systematic errors.
This approach has been intractable due to the large computational costs
on conventional CPU clusters, but rapid advances in accelerator
technology like GPUs open the possibility of simulation-based modeling
and inference \citep{CranmerEtAl2020}.
Furthermore, model differentiability enabled by AD libraries can
accelerate parameter constraint with gradient-based optimization and
inference.
A differentiable field-level forward model combining these two features
is able to constrain physical parameters together with the initial
conditions of the Universe.

The first differentiable cosmological simulations, such as BORG, ELUCID,
and BORG-PM \citep{BORG, ELUCID, BORG-PM}, were developed before the
advent of modern AD systems, and were based on \HL{the analytic derivatives,
which involve first convoluted derivation by hand using the chain rule
\citep[see e.g.,][App.~D]{SeljakEtAl2017} before implementing them in
code.}
Later codes including \texttt{FastPM} and \texttt{FlowPM} \citep{FastPM,
vmad, SeljakEtAl2017, FlowPM} compute gradients using the AD engines,
namely \texttt{vmad} (written by the same authors) and
\texttt{TensorFlow}, respectively.
\HL{The AD frameworks automatically apply the chain rule to the primitive
operations that comprise the whole simulation, relieving the burden of
derivation and implementation of the derivatives.}
Both analytic differentiation and AD backpropagate the gradients through
the whole history, which requires saving the states at all time steps in
memory.
Therefore, they are subject to a trade-off between time and space/mass
resolution, usually sacrificing both.
As a result, they lose accuracy on small scales and in dense regions
where the time resolution is important, e.g., in weak lensing
\citep{MADLens}.

\HL{Alternatively, the adjoint method provides systematic ways of deriving
the gradients of an objective function $\cJ$ under constraints
\citep{Pontryagin1962}, such as those imposed by the $N$-body equations
of motion in a simulated Universe.
It identifies a set of \emph{adjoint variables} $\vlambda$, dual to the
state variables $\vz$ of the model, and carrying the gradient
information of the objective function with respect to the model state
$\p\cJ / \p\vz$.
For time-dependent problems, the adjoint variables evolve
\emph{backward} in time by a set of equations dual to that of the
forward evolution, known as the \emph{adjoint equations}.
For continuous time, the adjoint equations are a set of differential
equations, while in the discrete case they become difference equations
which are practically a systematic way to structure the chain rule or
backpropagation.
Their initial conditions are set by the explicit dependence of the
objective on the simulation state, e.g., $\vlambda_n = \p\cJ / \p\vz_n$
if $\cJ$ is a function of the final state $\vz_n$.
Solving the adjoint equations can help us to propagate the gradient
information via the adjoint variables to the input parameters $\vtheta$,
to compute the objective gradients $\d\cJ / \d\vtheta$.
And we will see later that the propagated and accumulated gradients on
parameters come naturally from multiple origins, each reflecting a
$\vtheta$-dependence at one stage of the modeling in
\autoref{fig:model}.}

The backward adjoint evolutions depend on the states in the forward run,
which we can re-simulate with reverse time integration if the dynamics
is reversible, thereby dramatically reducing the memory cost
\citep{NeuralODE}.
Furthermore, we \HL{derive the discrete adjoint equations dual to the
discrete forward time integration, known as the discretize-then-optimize
approach \citep[e.g.,][]{ANODE}, to ensure gradients propagate backward
along the \emph{same} trajectory as taken by the forward time
integration.
This is in contrast with the optimize-then-discretize approach, that
numerically integrates the continuous adjoint equations, and is prone to
larger deviations between the forward and the backward trajectories due
to different discretizations \citep{LanzieriLanusseEtAl2022}.}
In brief, to compute the gradient we only need to evolve a simulation
forward, and then backward jointly with its dual adjoint equations.
We introduce the adjoint method first for generic time-dependent
problems in both continuous and discrete cases in \autoref{sec:bwd}, and
then present its application on cosmological simulation in
\autoref{sec:adj}.

@@

modes $\vomega$ indicate dependence on $\vtheta$ only, while double

@@

We describe different operators in \autoref{sec:fwd}: Boltzmann solver
(``Boltz'') and initial condition generator (``IC'') in

@@

\vspace{1em}
\section{Forward Simulation}
\label{sec:fwd}

We first review and formulate all components of the $N$-body
simulation-based forward model of the cosmological structure formation.

@@

The initial conditions of particles can be set perturbatively when
$|\vnabla \cdot \vs|$, the linear approximation of the density
fluctuation, is much less than 1.
We compute the initial displacements and momenta using the second order
Lagrangian perturbation theory \citep[2LPT,][]{BouchetEtAl1995}:
%
\begin{align}
\vs &= D_1 \vs^{(1)} + D_2 \vs^{(2)}, \nonumber\\
\vp &= a^2 \dot\vx = a^2 \dot\vs
  = a^2 H \bigl( D_1' \vs^{(1)} + D_2' \vs^{(2)} \bigr),
\label{lpt}
\end{align}
%
where $\vp$ is the canonical momentum\footnote{We omit the particle mass
$m$ in the canonical momentum $\vp = m a^2 \dot\vx$ throughout for
brevity.} for canonical coordinate $\vx$.
The temporal and spatial dependences separate at each order: the $i$-th
order growth factor $D_i$ is only a function of scale factor $a$ (or
time $t$), and the $i$-th order displacement field $\vs^{(i)}$ depends
only on $\vq$.
We have also used two types of time derivatives, $\dot\Box \triangleq
\d\Box / \d t$ and $\Box' \triangleq \d\Box / \d \ln a$, related by
$\dot\Box = H \Box'$.

Both the first and second order displacements are potential flows,
%
\begin{equation}
\vs^{(1)} = - \vnabla_\vq \phi_\vs^{(1)},
\quad
\vs^{(2)} = - \vnabla_\vq \phi_\vs^{(2)},
\end{equation}
%
with the scalar potentials sourced by
%
\begin{align}
\nabla_\vq^2 \phi_\vs^{(1)} &= \delta^{(1)}(\vq), \nonumber\\
%
\nabla_\vq^2 \phi_\vs^{(2)} &= {\textstyle \sum_{i<j}} \bigl(
  \phi_{\vs,ii}^{(1)}\phi_{\vs,jj}^{(1)}
  - \phi_{\vs,ij}^{(1)}\phi_{\vs,ji}^{(1)}
\bigr),
\end{align}
%
where $\phi_{\vs,ij} \triangleq \p^2\phi_\vs / \p q_i \p q_j$,
and $\delta^{(1)}$ is the linear order of the overdensity field,
$\delta$, related to the density field $\rho$ and mean matter density
$\bar\rho$ by $\delta \triangleq \rho / \bar\rho - 1$.

@@

Cosmological perturbation theory gives the linear power spectrum
as\footnote{Here the time dependence of the linear power, $\Plin(k, a) =
\Plin(k) D_1^2(a)$, has been left to \eqref{lpt}}
%
\begin{equation}
\frac{k^3}{2\pi^2} \Plin(k)
= \frac{4}{25} \As
  \Bigl( \frac{k}{k_\mathrm{pivot}} \Bigr)^{\ns - 1}
  T^2(k) \frac{c^4 k^4}{\Omegam^2 H_0^4},
\label{Plin}
\end{equation}
%
where the shape of $\Plin$ is determined by the transfer function $T$,
solution to the linearized Einstein-Boltzmann equations \citep{CAMB,
CLASS}.
$T$ depends on the cosmological parameters,
%
\begin{equation*}
\vtheta = (
  \As, \ns, \Omegam, \Omegab, h, \cdots),
\end{equation*}
%
some of which already appear in \eqref{Plin}: $\As$ is the amplitude of
the primordial power spectrum defined at some fixed scale
$k_\mathrm{pivot}$; $\ns$ describes the shape of the primordial power
spectrum; $\Omegam$ is the total matter density parameter; $\Omegab$ is
the baryonic matter density parameter; and $H_0$ is the Hubble constant,
often parameterized by the dimensionless $h$ as $H_0 = 100 h \,
\mathrm{km}/\mathrm{s}/\Mpc$.
Other parameters may enter in extensions of the standard $\Lambda$ cold
dark matter ($\Lambda$CDM) cosmology.

@@

While our adjoint method is general, we employ the PM solver in \pmwd\
for efficiency, and leave the implementation of short-range forces to
future development.
With the PM method, we evaluate $\delta(\vx)$ on an auxiliary mesh by
\emph{scattering} particle masses to the nearest grid points.
We use the usual cloud-in-cell (CIC), or trilinear, interpolation
\citep{HockneyEastwood1988}, to compute the fractions of a particle at
$\vx'$ going to a grid point at $\vx$,
%
\begin{equation}
W(\vx, \vx') = \prod_{i=1}^3
  \max \Bigl( 1 - \frac{|x_i-x'_i|}{l}, 0 \Bigr),
\label{cic}
\end{equation}
%
where $l$ is the mesh cell size.

@@

$N$-body particles move by the following equations of motion
%
\begin{align}
\dot\vx &= \frac{\vp}{a^2}, \nonumber\\
\dot\vp &= -\vnabla\phi = - \frac{\vnabla\varphi}a.
\label{eom}
\end{align}
%
We use the \texttt{FastPM} time stepping \citep{FastPM}, designed to
reproduce in the linear regime the linear Lagrangian perturbation
theory, i.e., the 1LPT as the first order in \eqref{lpt}, also known as
the the Zel'dovich approximation (hereafter ZA).
We present a simplified derivation below.

@@

\texttt{FastPM} chooses $G_D$ and $G_K$ according to the ZA growth
history, thereby improving the accuracy on large scales and at early
times.
In ZA, the displacements are proportional to the linear growth factor,
$\vs \propto D_1$, which determines the time dependences of the momenta
and the accelerations by \eqref{eom}.
Therefore, we can set $G_D$ and $G_K$ in \eqref{eom_integ} to
%
\begin{align}
G_D &:= a^2 \dot D_1, \nonumber\\
G_K &:= a \dot G_D.
\label{G_ZA}
\end{align}
%
They are functions of $D_1$ and its derivatives, as given by
\eqref{G_eval}.
With these choices, the drift and kick factors, defined in
\eqref{eom_integ}, then become
%
\begin{align}
D(t_c, t_a, t_b) &=
  \frac{D_1(t_b) - D_1(t_a)}{G_D(t_c)}, \nonumber\\
%
K(t_c, t_a, t_b) &=
  \frac{G_D(t_b) - G_D(t_a)}{G_K(t_c)}.
\end{align}

@@

\vspace{1em}
\section{Backward Differentiation -- the Adjoint Method}
\label{sec:bwd}

We first introduce the adjoint method for generic time-dependent ODEs,
and derive the adjoint equations following a pedagogical tutorial by
\citet{Bradley2019}.
We then adopt the discretize-then-optimize approach, and derive the
discrete adjoint equations, that is more suitable for the $N$-body
symplectic time integration.
Finally we apply them to derive the adjoint equations and the gradients
for cosmological simulations described in \autoref{sec:fwd}, and couple
them with the reverse time integration to reduce the space complexity.
